---
title: "432Project"
output: pdf_document
date: "2024-11-26"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

**0, descriptive statistics, missing values, outliers**

```{r}
data <- read.csv("diabetes_012_health_indicators_BRFSS2015.csv")
```

```{r}
library(dplyr)
data_bad_MH <- data |> 
  filter(MentHlth > 15) # Focus on groups with poor mental health
```

```{r}
# Check for missing values in each column
missing_values <- sapply(data_bad_MH, function(x) anyNA(x))
missing_values
```

There is no missing values

summary for non 0/1 variables: 

```{r}
summary(data_bad_MH[,c(  "BMI","MentHlth","PhysHlth","Age","Education","Income")])
```

Consider BMI for outliers: 

```{r}
hist(data_bad_MH$BMI, 
     breaks = 30, 
     col = "lightblue", 
     border = "black",
     main = "Distribution of BMI", 
     xlab = "BMI", 
     ylab = "Frequency",
     cex.main = 1.5, 
     cex.lab = 1.2, 
     cex.axis = 1)
```

```{r}
# save a version not inversed
diabete_data <- data_bad_MH
```

**1: Check correlations**

Reverse variables 0/1 to align: 1--"bad"  0--"good", except sex, age, edu, income

```{r}
# for these vars, 0 -> 1, 1 -> 0 to align healthy/unhealthy implications
reversed_vars <- c("PhysActivity", "Fruits", "Veggies", "CholCheck", "AnyHealthcare")
data_bad_MH[reversed_vars] <- lapply(data_bad_MH[reversed_vars], function(x) 1 - x)
```

Reversed variables: 

CholCheck:
1 = no cholesterol check in 5 years; 0 = yes cholesterol check in 5 years

PhysActivity:
physical activity in past 30 days - not including job 1 = no 0 = yes

Fruits
Consume Fruit 1 or more times per day 1 = no 0 = yes

Veggies
Consume Vegetables 1 or more times per day 1 = no 0 = yes

AnyHealthcare
Have any kind of health care coverage, including health insurance, prepaid plans such as HMO, etc. 1 = no 0 = yes

```{r}
library(corrplot)
cor_matrix <- cor(data[, setdiff(names(data), c("Sex", "Age", "Education", "Income"))])

corrplot(cor_matrix, 
         method = "color", 
         type = "upper", 
         tl.col = "black", 
         tl.srt = 45,  
         tl.cex = 0.8)
```

```{r}
excluded_vars <- c("Sex", "Age", "Education", "Income")
data_for_corr <- data_bad_MH[, !(names(data_bad_MH) %in% excluded_vars)]

# Calculate the correlation matrix
cor_matrix <- cor(data_for_corr)

# Print the correlation matrix
print(cor_matrix)
```

```{r}
threshold <- 0.3
large_corr <- which(abs(cor_matrix) > threshold, arr.ind = TRUE)

large_corr <- large_corr[large_corr[, 1] != large_corr[, 2], ]
large_corr <- large_corr[!duplicated(t(apply(large_corr, 1, sort))), ]

# Ones with large correlations

large_corr_values <- data.frame(
  Variable1 = rownames(cor_matrix)[large_corr[, 1]],
  Variable2 = colnames(cor_matrix)[large_corr[, 2]],
  Correlation = cor_matrix[large_corr]
)
print(large_corr_values)
```

Only moderate correlations, General Health large correlations make sense. Difficulty Walking and Physical Health also makes sense, and one is self-report the other is 0/1. We can keep all. (This is for if we didn't filter by mental health)

Consider removing PhysHlth or GenHlth

**2. VIF**

```{r}
library(car)

# Try different response variables for health: 
vif_model1 <- lm(Diabetes_012 ~ ., data = data_bad_MH)
vif_model2 <- lm(BMI ~ ., data = data_bad_MH)
vif_model3 <- lm(GenHlth ~ ., data = data_bad_MH)

# Calculate VIF values for each predictor
vif_values1 <- vif(vif_model1)
vif_values2 <- vif(vif_model2)
vif_values3 <- vif(vif_model3)

vif_values1
vif_values2
vif_values3

any(vif_values1 > 2|vif_values2 > 2|vif_values3 > 2)
```

Collinearity is not an issue. 


**3. Factor Analysis (Optional)**

```{r}
library(psych)

excluded_vars <- c("Diabetes_012", "Sex", "Age", "Education", "Income")
data_for_fa <- data_bad_MH[, !(names(data_bad_MH) %in% excluded_vars)]

# Check suitability of the data for factor analysis
# Bartlett’s test and KMO test
cortest.bartlett(data_for_fa) # Bartlett's test for sphericity
KMO(data_for_fa)             # Kaiser-Meyer-Olkin measure of sampling adequacy

# Determine the number of factors using a scree plot
fa.parallel(data_for_fa, fa = "fa", n.iter = 100) # Parallel analysis

# Perform Exploratory Factor Analysis (EFA)
# Choose the number of factors based on parallel analysis or eigenvalues > 1
num_factors <- 3 # Replace this with the chosen number of factors
efa_result <- fa(data_for_fa, nfactors = num_factors, rotate = "varimax")

# View factor loadings
print(efa_result$loadings, cutoff = 0.3) # Loadings above 0.3 are highlighted

# Visualize the factor structure
fa.diagram(efa_result)
```

```{r}
factor_scores <- factor.scores(data_for_fa, efa_result, method = "regression")

scores <- factor_scores$scores

# Add the factor scores back to the original dataset
data_bad_MH <- cbind(data_bad_MH, scores)

head(scores)
```

```{r}
# Extract the factor loadings
factor_loadings <- efa_result$loadings

# Convert factor loadings to a data frame for easier handling
factor_loadings_df <- as.data.frame.matrix(factor_loadings)

# Set a cutoff to determine significant loadings (e.g., 0.3)
loading_cutoff <- 0.3

# Identify variables associated with each factor
MR1_vars <- rownames(factor_loadings_df[abs(factor_loadings_df[, "MR1"]) > loading_cutoff, ])
MR2_vars <- rownames(factor_loadings_df[abs(factor_loadings_df[, "MR2"]) > loading_cutoff, ])
MR3_vars <- rownames(factor_loadings_df[abs(factor_loadings_df[, "MR3"]) > loading_cutoff, ])

# Print the variables associated with each factor
cat("Variables for MR1 (Health indicators, self-reports):\n", paste(MR1_vars, collapse = ", "), "\n\n")
cat("Variables for MR2 (Health Behaviors/Diet):\n", paste(MR2_vars, collapse = ", "), "\n\n")
cat("Variables for MR3 (Healthcare):\n", paste(MR3_vars, collapse = ", "), "\n\n")
```

filter:
1. Low mental health + no healthcare 

identifier: 
2. Education + Income + ... (other variables not in model)


```{r}
data_no_HC <- data |> 
  filter(AnyHealthcare == 1) |> 
  select(-AnyHealthcare)# Compare with having HC

data_HC <- data |> 
  filter(AnyHealthcare == 0) |> 
  select(-AnyHealthcare)
```


Repeat for data_no_HC: 

```{r}
library(psych)

excluded_vars <- c("Diabetes_012", "Sex", "Age", "Education", "Income")
data_for_fa_no_HC <- data_no_HC[, !(names(data_no_HC) %in% excluded_vars)]

# Check suitability of the data for factor analysis
# Bartlett’s test and KMO test
cortest.bartlett(data_for_fa_no_HC) # Bartlett's test for sphericity
KMO(data_for_fa_no_HC)             # Kaiser-Meyer-Olkin measure of sampling adequacy

# Determine the number of factors using a scree plot
fa.parallel(data_for_fa_no_HC, fa = "fa", n.iter = 100) # Parallel analysis

# Perform Exploratory Factor Analysis (EFA)
# Choose the number of factors based on parallel analysis or eigenvalues > 1
num_factors <- 3 # Replace this with the chosen number of factors
efa_result_no_HC <- fa(data_for_fa_no_HC, nfactors = num_factors, rotate = "varimax")

# View factor loadings
print(efa_result_no_HC$loadings, cutoff = 0.3) # Loadings above 0.3 are highlighted

# Visualize the factor structure
fa.diagram(efa_result_no_HC)
```

```{r}
# Extract the factor loadings
factor_loadings_no_HC <- efa_result_no_HC$loadings

# Convert factor loadings to a data frame for easier handling
factor_loadings_df_no_HC <- as.data.frame.matrix(factor_loadings_no_HC)

# Set a cutoff to determine significant loadings (e.g., 0.3)
loading_cutoff_no_HC <- 0.4

# Identify variables associated with each factor
MR1_vars_no_HC <- rownames(factor_loadings_df_no_HC[abs(factor_loadings_df_no_HC[, "MR1"]) > loading_cutoff_no_HC, ])
MR2_vars_no_HC <- rownames(factor_loadings_df_no_HC[abs(factor_loadings_df_no_HC[, "MR2"]) > loading_cutoff_no_HC, ])
MR3_vars_no_HC <- rownames(factor_loadings_df_no_HC[abs(factor_loadings_df_no_HC[, "MR3"]) > loading_cutoff_no_HC, ])

# Print the variables associated with each factor
cat("Variables for MR1_no_HC (Health indicators, self-reports):\n", paste(MR1_vars_no_HC, collapse = ", "), "\n\n")
cat("Variables for MR2_no_HC (Health Behaviors/Diet):\n", paste(MR2_vars_no_HC, collapse = ", "), "\n\n")
cat("Variables for MR3_no_HC (Healthcare):\n", paste(MR3_vars_no_HC, collapse = ", "), "\n\n")
```


Repeat for data_HC


```{r}
library(psych)

excluded_vars <- c("Diabetes_012", "Sex", "Age", "Education", "Income")
data_for_fa_HC <- data_HC[, !(names(data_HC) %in% excluded_vars)]

# Check suitability of the data for factor analysis
# Bartlett’s test and KMO test
cortest.bartlett(data_for_fa_HC) # Bartlett's test for sphericity
KMO(data_for_fa_HC)             # Kaiser-Meyer-Olkin measure of sampling adequacy

# Determine the number of factors using a scree plot
fa.parallel(data_for_fa_HC, fa = "fa", n.iter = 100) # Parallel analysis

# Perform Exploratory Factor Analysis (EFA)
# Choose the number of factors based on parallel analysis or eigenvalues > 1
num_factors_HC <- 3 # Replace this with the chosen number of factors
efa_result_HC <- fa(data_for_fa_HC, nfactors = num_factors_HC, rotate = "varimax")

# View factor loadings
print(efa_result_HC$loadings, cutoff = 0.3) # Loadings above 0.3 are highlighted

# Visualize the factor structure
fa.diagram(efa_result_HC)
```


```{r}
# Extract the factor loadings
factor_loadings_HC <- efa_result_HC$loadings

# Convert factor loadings to a data frame for easier handling
factor_loadings_df_HC <- as.data.frame.matrix(factor_loadings_HC)

# Set a cutoff to determine significant loadings (e.g., 0.3)
loading_cutoff_HC <- 0.4

# Identify variables associated with each factor
MR1_vars_HC <- rownames(factor_loadings_df_HC[abs(factor_loadings_df_HC[, "MR1"]) > loading_cutoff_no_HC, ])
MR2_vars_HC <- rownames(factor_loadings_df_HC[abs(factor_loadings_df_HC[, "MR2"]) > loading_cutoff_no_HC, ])
MR3_vars_HC <- rownames(factor_loadings_df_HC[abs(factor_loadings_df_HC[, "MR3"]) > loading_cutoff_no_HC, ])

# Print the variables associated with each factor
cat("Variables for MR1_HC (Health indicators, self-reports):\n", paste(MR1_vars_HC, collapse = ", "), "\n\n")
cat("Variables for MR2_HC (Health Behaviors/Diet):\n", paste(MR2_vars_HC, collapse = ", "), "\n\n")
cat("Variables for MR3_HC (Healthcare):\n", paste(MR3_vars_HC, collapse = ", "), "\n\n")
```

